---
title: "Prediction of NBA Player's Salary Based on Stepwise Regression Model"
author: "Hongye Xu"
output:
  html_document:
    df_print: paged
  pdf_document:
    keep_tex: yes
    latex_engine: xelatex
---
---
#1 Introduction
## 1.1 Background
I have always felt that the NBA has the best data storage in the sport filed. In the beginning, I wanted to analyze the performance of the players by scrapping the data from the official NBA.stat website. However, since the NBA.stat table is in javascript format, and the official has canceled all the existing official APIs, no possible R-based crawler method has been found after the effort. Therefore, I chose an alternative, which is the basketball-reference website. This report is based on two data sources on the basketball-reference. My goal is to predict the player's salary for next season based on player performance this season.

## 1.2 Glossary
Abbreviation | Explanation
------------- | -------------
Pos | Position
Age | Age of Player at the start of February 1st of that season 
Tm | Team 
G | Games
GS | Games Started 
MP | Minutes Played Per Game 
FG | Field Goals Per Game
FGA | Field Goal Attempts Per Game
FG% | Field Goal Percentage 
3P | 3-Point Field Goals Per Game
3PA | 3-Point Field Goal Attempts Per Game
3P% | FG% on 3-Pt FGAs.
2P | 2-Point Field Goals Per Game 
2PA | 2-Point Field Goal Attempts Per Game 
eFG% | Effective Field Goal Percentage
FT | Free Throws Per Game
FTA | Free Throw Attempts Per Game 
FT% | Free Throw Percentage 
ORB | Offensive Rebounds Per Game 
DRB | Defensive Rebounds Per Game
TRB | Total Rebounds Per Game 
AST | Assists Per Game 
STL | Steals Per Game 
BLK | Blocks Per Game 
TOV | Turnovers Per Game 
PF | Personal Fouls Per Game 
PTS | Points Per Game 

#2 Preparation
```{r,echo=FALSE,warning=FALSE,message=FALSE}
library(rvest)
library(magrittr)
library(tibble)
library(dplyr)
library(stringr)
library(data.table)
library(corrplot)
library(GGally)
library(tidyverse)
library(PerformanceAnalytics)
library(plotly)
library(caret)
library(MASS)
library(knitr)
library(png)
```

##2.1 Data Scraping and Cleaning 
###2.1.1 Players' Regular Season Data
My first data source came from https://www.basketball-reference.com/leagues/NBA_2019_per_game.html. Because all the data in the website is displayed in the form of HTML table, I can read the data in this table by reading the XPath('//*[@id="per_game_stats"]') of the table with Chrome browser.
At the same time, because many NBA players transfer during the season, this data source records their data in many teams, and we just need to keep the data that represents their season average.In addition, I deleted those rows which do not contain 3 points field goals, 2 points field goals and free throw field goals.
```{r echo=FALSE}
#NBA Season18-19 Players stat(Regular Season)
url<- "https://www.basketball-reference.com/leagues/NBA_2019_per_game.html"
player_season_v1 <- url %>% read_html()%>%
  html_nodes(xpath = '//*[@id="per_game_stats"]')%>%
  html_table()
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# romove the rank column and row number
player_season_v2 <- player_season_v1[[1]]%>%dplyr::select(-Rk)
player_season_v3 <-player_season_v2[- grep("Player", player_season_v2$Player),]
# make some columns numeric
name <- c("Age","G","GS","MP","FG","FGA","FG%","3P","3PA","3P%","2P","2PA","2P%","eFG%","FT","FTA","FT%","ORB","DRB","TRB","AST","STL","BLK","TOV","PF","PTS")
player_season_v3[name] <- sapply(player_season_v3[name],as.numeric)
rownames(player_season_v3) <- NULL

# There will be multiple rows of data for players transferred during the season. We only keep the one with the largest number of games played(which is the stat of this player throughout the season).
player_season_tidy <- player_season_v3 %>% group_by(Player)%>%
   mutate(rank = min_rank(desc(G))) %>%
   filter(rank == 1) %>%
   dplyr::select(-rank)
# No Scale data(for data visulization)
player_season_tidy <- player_season_tidy %>% filter(!is.na(`3P%`) & !is.na(`FT%`)& !is.na(`2P%`)) %>% as.data.frame(.)
head(player_season_tidy)
```

### 2.1.2 Scale
Considering that regression analysis is mainly used in this report. In order to eliminate the inaccuracy of parameters caused by too large or too small data itself. I create a standardized version of the data.
```{r,echo=FALSE}
#Here i use the Player column as the rowname since the scale function need the whole matrix features to be numeric
scale <- player_season_tidy %>% dplyr::select(Player ,Age,MP,`2P%`,`3P%`,`FT%`,TRB:PTS)
rownames(scale)<-scale[,1]
scale1 <- scale[,-1]%>% as.matrix(.)%>%
  scale(.) %>%
  as.data.frame(.)
head(scale1)
```

###2.1.3 Players' Salaries
The second data came from https://www.basketball-reference.com/contracts/players.html, and I imported it in the same way as I scrapped the first data. This data includes salary data for the next six seasons. But all we need is this season's data, so I filtered it and converted character data into numeric data. The result is as follow.
```{r,echo=FALSE}
# Players' salaries from season 18-19
url1 <- "https://www.basketball-reference.com/contracts/players.html"
salaries <- url1 %>% read_html()%>%
  html_nodes(xpath = '//*[@id="player-contracts"]')%>%
  html_table()
salaries_v2 <- salaries[[1]]
colnames(salaries_v2) <- NULL
names(salaries_v2) <- as.character(unlist(salaries_v2[1,]))
salaries_v2 <- salaries_v2[-1,] %>%
  dplyr::select(-Rk)
rownames(salaries_v2) <- NULL
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
# Change 2018-19 salaries to be numeric data
salaries_v2$`2018-19` <- salaries_v2$`2018-19` %>%
  str_replace_all(.,"\\,","")%>%
  str_replace_all(., "\\$","")%>%
  as.numeric(.)
# Delete rows containing missing values (due to duplicate headers)
salaries_v2 <-salaries_v2[- grep("Player", player_season_v2$Player),]
salaries_v3 <-na.omit(salaries_v2)
#remove duplicated row(only keep the highest one)
salaries_v4 <- salaries_v3 %>% group_by(Player) %>%
  mutate(rank = min_rank(desc(`2018-19`))) %>%
   filter(rank == 1) %>%
   dplyr::select(-rank)
```

```{r,echo=FALSE,warning=FALSE,message=FALSE}
salaries_tidy <- salaries_v4 %>% 
  dplyr::select(Player,Tm,`2018-19`)%>%
  as.data.frame(.)
# Following two players' salaries are not changed after transfer(So i delete them to avoid duplication of one player after doing table merging)
salaries_tidy <- salaries_tidy[!(salaries_tidy$Player=="John Jenkins" & salaries_tidy$Tm=="NYK"),]
salaries_tidy <- salaries_tidy[!(salaries_tidy$Player=="Emanuel Terry" & salaries_tidy$Tm=="MIA"),]
salaries_tidy %>% mutate(duplicated(Player))%>% 
  filter(`duplicated(Player)`== TRUE)
head(salaries_tidy) 
```
Finally, there is no duplicate player data.


###2.1.4 Merging Data
Merge standardized and non-standardized data with player salary data to get the final data we need.
```{r,echo=FALSE}
#non_scale data
stats_for_visualized <- merge(player_season_tidy, salaries_tidy, by.x = "Player", by.y = "Player")
names(stats_for_visualized)[31] <- "salary18_19"
stats_for_visualization <- stats_for_visualized[-30]
```

```{r,echo=FALSE}
#scale data
salaries1 <- salaries_tidy 
rownames(salaries1) <- salaries1[,1]
salaries2 <- salaries1[,-1]
stats_scale <- merge(scale1, salaries2, by="row.names")
names(stats_scale)[15] <- "salary18_19"
stats_scale <- stats_scale[-14]
```
###2.1.5 Save No_scale Data Into CSV
```{r,echo=FALSE}
write.csv(stats_for_visualization,'18-19players_stat.csv')
```
I saved the merged non-standardized data to the working directory with the file name'18-19 players_stat.csv'.

# 3 Correlation Check
## 3.1 Frist Check
```{r,echo=FALSE}
corrplot(cor(stats_scale %>% 
               dplyr::select(salary18_19, Age:PTS, contains("%")), 
             use = "complete.obs"), 
         method = "circle",type = "upper")
```

The features that have strong correlation with salary are:PTS,TOV,STL,AST,TRB and MP.
Besides, MP is strongly correlated with multiple features and may have multiple collinearities(This is in line with our common sense. The more time we play, the better the data will be). What I didn't expect was that the correlation between field goal and salary was not high, that is to say, the output of players influenced the salary of players more than efficiency.

## 3.2 Second Check
```{r,warning=FALSE,message=FALSE,echo=FALSE}
stats_salary_cor <- 
  stats_scale %>% 
  dplyr::select(salary18_19,PTS, MP, TOV, AST, STL, TRB)
ggpairs(stats_salary_cor)
cor(stats_salary_cor)[,"salary18_19"]
```


Correlation strength is: PTS > MP > TOV > AST > STL > TRB
There's also one thing that surprises me: the number of players'turnivers is positively correlated with their salaries. I mean, generally speaking, assuming that a player's turnover rate is constant, the total number of turnovers will increase as his minutes played increases, and important players will have higher minutes played and higher salaries.

#4 Data Visualization
##4.1 Interactive Plot
```{r,echo=FALSE,warning=FALSE,message=FALSE}
names(stats_for_visualization)[4] <- "Team"
# for html output-------
plot_ly( format = "pdf",data = stats_for_visualization , x = ~salary18_19, y = ~PTS, color = ~Team,
       hoverinfo = "text",
       text = ~paste("Player: ", Player,
                     "<br>Salary: ", format(salary18_19, big.mark = ","),"$",
                    "<br>PTS: ", round(PTS, digits = 3),
                     "<br>Team: ", Team)) %>% 
 layout(
   title = "Salary vs Points Per Game",
   xaxis = list(title = "Salary USD"),
   yaxis = list(title = "Points per Game")
 )
# for html output-------

# for pdf output ------
#a <- plot_ly( format = "pdf",data = stats_for_visualization , x = ~salary18_19, y = ~PTS, color = ~Team,
#       hoverinfo = "text",
#       text = ~paste("Player: ", Player,
#                     "<br>Salary: ", format(salary18_19, big.mark = ","),"$",
#                    "<br>PTS: ", round(PTS, digits = 3),
#                     "<br>Team: ", Team)) %>% 
# layout(
#   title = "Salary vs Points Per Game",
#   xaxis = list(title = "Salary USD"),
#   yaxis = list(title = "Points per Game")
# )
#tmpFile <- tempfile(fileext = ".png")
#export(a, file = tmpFile)
# for pdf output ------
```

##4.2 Scatter Plot With Regression Line
```{r,echo=FALSE}
stats_for_visualization %>% 
  ggplot(aes(x = salary18_19, y = PTS)) + 
  geom_point() + 
  geom_smooth(method = "lm") 
```

Under the simple linear model, we can understand that the fitted curve represents the average level of the league, and the player below the curve performs worse than the expected performance corresponding to the salary. We can check their name 
by hovering on the points in the interactive plot(only in HTML form file). It includes a lot of All-Star players, such as Chris Paul, Kyle Lowry, Al Horford, Gordon Haywood(The Celtics are unlucky) and etc. However, it only considers the scoring feature, and does not fully reflect the players'influence on the field.

#5 Multiple Regression
```{r,echo=FALSE}
regression <- stats_scale %>% dplyr::select(salary18_19,PTS, MP, TOV, AST, STL, TRB)
  lm(salary18_19~ PTS + MP + TOV + AST + STL + TRB, data =regression)
```
From here, we can see that points per game is the most significant feature of positive impact, while turnovers per game is the most significant feature of negative impact. However, simple multiple regression also has some problems, that is, there are multiple collinearities.

## 5.1 Player 's Importance And Incautiousness
Here we make two definitions that a player is "important" if his minutes played is above average and is "incautious" if his turnover per game is above average.

```{r,echo=FALSE}
avg.minutes <- mean(regression$MP)
avg.turnover <- mean(regression$TOV)
regression$Importance<- as.factor(ifelse(regression$MP >= avg.minutes, "Yes", "No"))
regression$Incautiousness <- as.factor(ifelse(regression$TOV >= avg.turnover, "Yes", "No"))
```

##5.2 Prallel Slope Model
###5.2.1 Incautiousness Comparision
```{r,echo=FALSE}
regression %>% 
  ggplot(aes(x = salary18_19, y = (TOV * var(player_season_tidy$TOV)^(1/2) + mean(player_season_tidy$TOV)), colour = Incautiousness)) + 
  geom_point() + 
  geom_smooth(method="lm")+
  ylab("turnovers per game")
```

The plot shows that the number of turnovers of the players with higher salaries will increase correspondingly, but the magnitude is not large. We can think that it is a natural phenomenon caused by the increase of playing time. 
Then I do a regression analysis of Importance and Incautiousness. The result is as follow
```{r,echo=FALSE}
lm(formula = salary18_19 ~ Importance * Incautiousness, data=regression)
```

We can assume that the impact of A and B is close to synchronization, which confirms my previous view that players with higher salaries have more playing time, which leads to more turnovers, rather than higher salaries because of higher turnovers.

## 5.3 Stepwise Regression
Considering that the data of NBA players will increase with the increase of playing time, there must be multiple collinearity among the features. So stepwise regression is the more accurate method.
```{r,echo=FALSE}
# Do Stepwise Regression
full.model <- lm(salary18_19~., data =regression)
step.model <- stepAIC(full.model, direction = "both", 
                      trace = FALSE)
summary(step.model)
```

Here I use k-fold cross-validation to test the error of the models that have different number of variate.
```{r,echo=FALSE}
# Set seed for reproducibility
set.seed(123)
# Set up repeated k-fold cross-validation
train.control <- trainControl(method = "cv", number = 10)
# Train the model
step.model <- train(salary18_19 ~., data = regression,
                    method = "leapBackward", 
                    tuneGrid = data.frame(nvmax = 1:9),
                    trControl = train.control
                    )
step.model$results
```
From the result we can see that three-variable model's RMSE is the smallest and Rsquared is second largest. So the three-variable model is the best one. Let us find out the order in which variables are added to the model.

```{r,echo=FALSE}
summary(step.model$finalModel)
coef(step.model$finalModel, 3)
```

The best model is salary18_19 ~ PTS + AST + TRB

# 6 Conclusion
## 6.1 What i want to predict
As Lebron James  fan, I am concerned about the new contract for the Lakers who may stay next season. Let's find them first.
```{r,echo=FALSE}
salaries_v4 %>% filter(Tm == "LAL" & `2019-20` == "")%>%
  dplyr::select(Player)
Pope <- stats_scale %>% filter(Row.names == "Kentavious Caldwell-Pope")
Bullock <- stats_scale %>% filter(Row.names == "Reggie Bullock")
Stephenson <- stats_scale %>% filter(Row.names == "Lance Stephenson")
```

## 6.2 Analysis conclusion
```{r,echo=FALSE}
salary_prediction <- function(m, point,assists,rebounds){
  pre_new <- predict(m, data.frame(PTS = point, AST= assists, TRB= rebounds))
  msg <- paste( "Expected Salary: $", format(round(pre_new), big.mark = ","), sep = "")
  print(msg)
}
model <- lm(salary18_19~ PTS + AST + TRB, data =regression)
salary_prediction(model,Pope$PTS,Pope$AST,Pope$TRB)
salary_prediction(model,Bullock$PTS,Bullock$AST,Bullock$TRB)
salary_prediction(model,Stephenson$PTS,Stephenson$AST,Stephenson$TRB)
```
Here I choose three players who are more likely to stay next season to make predictions.The result shows that salaries for Pope, Bullock and Stephenson for next season are $6,771,542, $7,275,901 and $5,902,181


#7 Github Link
https://github.com/szxuhongye/NBA-Player-Salary-Predicton.git




